Software development is a complex task that virtually always involves developers collaborating. In this sense, many Free Libre and Open Source Software (FLOSS) development models have successfully created large, scalable, and (more often than not) globally distributed communities that evolve and maintain high-quality software projects, with the Linux kernel ecosystem being a prime example. Beyond the technical challenges natural to the project, the entry barrier is steep and discouraging for newcomers to the Linux ecosystem as there are many sub-projects (called subsystems), each with its specific contributing rules, processes, and practices that are usually undocumented and only learned by direct contact with the community; this risks the project long-term sustainability, as the renewal of the highly qualified workforce is a known problem. This work aims to validate an approach to mentor newcomers to the Linux ecosystem efficiently. The research questions that guided us are: (1) What teaching techniques are effective in introducing newcomers to the Linux ecosystem (in-loco workshops, tutorials, lectures from practitioners)? (2) What hard and soft skills are crucial to successfully entering the Linux ecosystem? (3) What processes and practices (workflows) should the mentors abstract for a smoother entry into the Linux ecosystem? In the first semesters of 2024 and 2025, the Free Software Development course was ministered at the University of São Paulo (USP) by the authors - one as the professor and the others as teaching assistants (TAs). During these offerings, students went from setting up a testing environment to learning the fundamental workflows involved, culminating in sending patches and interacting with Linux communities through the code review process. These students, who in the majority had no experience in Linux development, were closely mentored using a combination of teaching techniques, and their experience and feedback were collected through surveys and blog posts written by them. Among our findings, we can highlight: (1) Use of directed content (tutorials) produced by real practitioners in the Linux ecosystem, along with in-person mentoring during classes (workshops) and accessibility of the professor and TAs, produces a fertile environment for newcomers; (2) The experience in the course enhanced the qualification of students, from hard skills like git, and email and web-based models of code collaboration to communication skills; (3) The experience in the course demystified a lot of inaccurate concepts from the students about FLOSS development and made them more comfortable and ready to contribute to other FLOSS projects.