\section{Introduction}
\label{sec:intro}

The development of modern software systems is a highly collaborative endeavor,
often demanding the coordination of multiple individuals with diverse skill
sets, experiences, backgrounds, and - more often than not - globally
distributed. Free Libre and Open Source Software (FLOSS) development models
employed by many successful software projects pervasive in our society have
raised the attention of industry and academia to leverage its benefits and get
insights into fields like Software Engineering (SE).

In these projects that employ FLOSS development models, contributions frequently
originate from a distributed and (sometimes) volunteer-based workforce called
contributors. At the same time, the approval, feedback, and decision-making of
the changes are the responsibility of a smaller group with administrative
privileges of the project. FLOSS development models have historically
demonstrated a unique ability to sustain long-lasting, high-quality software,
even under decentralized organization and asynchronous collaboration
constraints. The prime example of this success is the Linux kernel project,
which has grown into one of the largest and most influential FLOSS initiatives
in history. When we talk about the Linux project, we are talking about the
umbrella project composed of many sub-projects (called subsystems) that form a
complex and interconnected ecosystem.

Even though the Linux ecosystem is often celebrated for its technical excellence
and community-driven development, some challenges risk the project's long-term
sustainability. In particular, we want to focus on the problem of the steep
entry barrier to the Linux ecosystem for newcomers, which can be intimidating,
if not prohibitive, in some cases. Prospective contributors must navigate an
enormous codebase fragmented in many development contexts and dedicated code
repositories. These comprise the numerous subsystems governed by distinct and
often undocumented conventions, processes and practices (workflows), and social
norms. These characteristics complicate the onboarding process and pose a
problem for renewing the highly specialized workforce. Without a steady influx
of new and adequately trained contributors, the vitality and evolution of the
ecosystem could be jeopardized.

Addressing this issue requires more than technical tutorials or improved
documentation. It demands a pedagogical approach that recognizes the interplay
of technical proficiency, social integration, and community practices. In this
context, mentorship is necessary to keep newcomers motivated and smooth their
entry into the ecosystem. Effective mentorship can help demystify the
contribution process, transmit tacit knowledge, and build confidence in
potential contributors.

This paper presents and evaluates a structured approach to mentoring newcomers
to the Linux kernel ecosystem. Our research is grounded in practical experience
teaching the Free Software Development course at the University of São Paulo
(USP) during the first semesters of 2024 and 2025. The course, taught by the
authors - one as the professor and the others as teaching assistants (TAs) -
offered students a hands-on experience in FLOSS contribution focused on the
Linux project. The students learned to set up a testing environment; configure,
compile, and install custom-built kernels and modules; assess contribution
opportunities; develop and send contributions to maintainers and mailing lists;
and participate in the review process by interacting with the suggestions and
requests of the community to refine initially proposed changes to the standard
the project demands.

Through classroom workshops, curated tutorials authored by veteran Linux
practitioners, and sustained direct mentoring (by the professor and TAs),
students (in the majority with no experience with Linux or FLOSS development in
general) went from learning the basics to (in most cases) having a merged
contribution, within a couple of months. Qualitative data from students was
collected via surveys and through analyzing blog posts to examine their learning
trajectories and perceptions. Observations from the professor and TAs were also
systematically compiled to enrich the research.

The findings from this work indicate many interesting perspectives, which we
highlight:

\begin{enumerate}
    \item In-loco workshops, learning materials produced by veteran Linux
    practitioners, and accessible mentors can significantly lower the entry
    barrier to FLOSS projects like the Linux kernel;
    \item The students, independent of their background, reported that they
    enhanced their technical skills related to software development (deeper
    proficiency in git, device drivers, and C programming language) while
    exercising their communication skills, which is paramount for successful
    collaborative development;
    \item The students who had detached and imprecise perceptions of Linux and
    FLOSS development, which was the majority, reported that the mentoring
    experience helped them demystify those concepts and that they felt more
    empowered and comfortable to contribute to other FLOSS projects.  
\end{enumerate}


These contributions help to have a more comprehensive view of the movement of
new contributors to the Linux ecosystem and provide an approach to successfully
mentor them that can be applied to FLOSS projects in general.